Hydra Battles and Cie

### 2021-06-04 doc/part-hydras.tex
update Definition to On_Generic

\begin{itemize}
\item A ``mathematical'' representation of the set of countable ordinal numbers, afer Kurt Schütte~\cite{schutte}. This representation uses several (hopefully harmless) axioms. We use it as a reference for proving the correctness of ordinal notations.
\item A family of \emph{ordinal notations}, \emph{i.e.} data types used to represent segments $[0,\mu)$, where $\mu$ is some countable ordinal. Each ordinal notation is defined inside the Calculus of Inductive Constructions (without axioms). Many functions are defined, allowing proofs by computation. Note that proofs of 
correctness of a given ordinal notation with respect to Schütte's model obviously use axioms.
Please execute the \texttt{Print Assumptions} command in case of doubt.
\end{itemize}

@@

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as quite simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{[ordinal] notation systems}).

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}}

@@

Notation "'omega'" := (inr  0:t).
\end{Coqsrc}

@@

Notation "'omega'" := (1,0) : ON_scope.

@@

\begin{figure}[h]
   \centering
   \begin{tikzpicture}[very thick, scale=0.6]
\begin{scope}[color=blue]
\node (A) at (0,0) {$A$};
\node(A0) at (2,0)[label=below:$0$]{$\bullet$};
\node(A1) at (3,0)[label=below:$1$]{$\bullet$};
\node(A2) at (4,0)[label=below:$2$]{$\bullet$};
\node (Adots) at (6,0) {$\ldots$};
\end{scope}
\begin{scope}[color=red]
\node (B) at (0,2) {$B$};
\node(B0) at (2,2)[label=above:$0$]{$\bullet$};
\node(B1) at (3,2)[label=above:$1$]{$\bullet$};
\node(B2) at (4,2)[label=above:$2$]{$\bullet$};
\node (Bdots) at (6,2) {$\ldots$};
\node (b) at (8,2) [label=above:$\alpha$]{$\bullet$};
\node (bsucc) at (9,2) [label=above:$\alpha+1$]{$\bullet$};
\node (Bdots2) at (10,2) {$\ldots$};
\end{scope}
\begin{scope}[color=red!50!blue]
\draw [->,thin] (A0) -- node [auto] {$\iota$} (B0);
\draw [->,thin] (A1) -- node [auto] {$\iota$} (B1);
\draw [->,thin] (A2) -- node [auto] {$\iota$} (B2);
\draw [->,thin] (Adots) -- node [auto] {$\iota$} (Bdots);
\end{scope}
\end{tikzpicture}
   \caption{\textcolor{blue}{$A$} is a sub-segment  of \textcolor{red}{$B$}}
   \label{fig:subsegment}
 \end{figure}

@@

\begin{Coqsrc}
Notation "'FS' n" :=
     (ocons zero n zero) (at level 10) : t1_scope.
\end{Coqsrc}

@@

\label{sect:omega-notation2}
\begin{Coqsrc}
Notation omega := (ocons (ocons zero 0 zero) 0 zero): t1_scope.
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive ppT1 : Set :=
    P_fin : nat -> ppT1
  | P_add : ppT1 -> ppT1 -> ppT1
  | P_mult : ppT1 -> nat -> ppT1
  | P_exp : ppT1 -> ppT1 -> ppT1
  | P_omega : ppT1
\end{Coqsrc}

@@

\begin{Coqanswer}
     = (omega ^ omega + omega ^ 3 * 5 + 2)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
Fixpoint compare (alpha alpha':T1):comparison :=
  match alpha, alpha' with
    zero, zero => Eq
  | zero, ocons a' n' b' => Lt
  | _   , zero => Gt
  | (ocons a n b),(ocons a' n' b') =>
      (match compare a a' with 
          | Lt => Lt
          | Gt => Gt
          | Eq => (match lt_eq_lt_dec n n'
                   with
                       inleft  (left _) => Lt
                     | inright _ => Gt
                     |   _ => compare b b'
                   end)
       end)
  end.
\end{Coqsrc}

@@

Notation "'omega'"  := _Omega : E0_scope.
\end{Coqsrc}

@@

@@

\index{hydras}{Projects}
\begin{project}
The class \texttt{ON} of ordinal notations has been defined long after this 
chapter, and is not used yet in the development of the type \texttt{E0}. 
A better integration of both notions should simplify the development on ordinals in Cantor normal form. This integration is planned for the future versions.

@@

\section{A Variant for hydra battles}

@@

Our definition may look more complex than the mathematical one, but
uses plain structural recursion over the type \coqsimple{T1}. Thus, tactics like
\coqsimple{cbn}, \coqsimple{simpl}, \coqsimple{compute}, etc., are applicable. For simplicity's sake, 
we use an auxiliary function \texttt{canonS} of type \texttt{T1 -> nat  -> T1} such that
(\texttt{canonS  $\alpha$ $i$}) is equal to $\canonseq{\alpha}{i+1}$.

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Canon.html\#canonS}{Epsilon0.Canon}}

\label{Functions:canonS}
\label{Functions:canon}
\begin{Coqsrc}
Fixpoint canonS alpha (i:nat) :=
  match alpha with
      zero => zero
    | ocons zero 0 zero => zero
    | ocons zero (S k) zero => FS k
    | ocons gamma 0 zero =>
      match pred gamma with
          Some gamma' => ocons gamma' i zero
        | None => ocons (canonS gamma i) 0 zero
      end
    | ocons gamma (S n) zero =>
       match pred gamma with
           Some gamma' => ocons gamma n (ocons gamma' i zero)
         | None => ocons gamma n (ocons (canonS gamma i) 0 zero)
       end
    | ocons alpha n beta => ocons alpha n (canonS beta i)
  end.
\end{Coqsrc}

The following function computes $\canonseq{\alpha}{i}$, except for the case $i=0$, where it simply returns $0$\;\footnote{This restriction did not prevent us from proving all the main theorems of~\cite{KS81, KP82}. Nevertheless, in a future version of this development, we may define $\canonseq{\alpha}{0}$ exactly as 
in~\cite{KS81}. But we are afraid this would  be done at the cost of making some proofs much more complex.}.

\begin{Coqsrc}
 Definition canon alpha i := 
   match i with 0 => zero | S j => canonS alpha j end. 
\end{Coqsrc}

@@

\begin{Coqsrc}
Compute (canon (omega ^ omega) 3).
\end{Coqsrc}

\begin{Coqanswer}
  = phi0 (FS 2) : T1
\end{Coqanswer}

\begin{Coqsrc}
Example canon3 :  canon (omega ^ omega) 3 = omega ^ 3.
Proof. reflexivity. Qed.
\end{Coqsrc}

\begin{Coqsrc}
Compute pp (canon (omega ^ omega * 3) 42).  
\end{Coqsrc}

\begin{Coqanswer}
    = (omega ^ omega * 2 + omega ^ 42)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma canonS_succ i alpha :
  nf alpha ->  canonS (succ alpha) i = alpha.
Proof.
 induction alpha.
 (* ... *)
\end{Coqsrc}

@@

\label{lemma:canonS_LT}
\begin{Coqsrc}
Lemma canonS_LT i alpha :
  nf alpha -> alpha <> zero -> canonS alpha i t1<  alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma canonS_limit_strong (lambda : T1) : 
     nf lambda ->
     limitb lambda  ->
     forall beta, beta t1< lambda ->
                  {i:nat | beta t1< canonS lambda i}.

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : Paths}
\index{maths}{Ordinal numbers!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\begin{Coqsrc}
(Definition transition_S i : relation T1 :=
  fun alpha beta =>  alpha <> zero /\ beta = canonS alpha i.

@@

\begin{Coqsrc}
Lemma canonS_iota i alpha :
    nf alpha -> alpha <> 0 ->
    iota alpha -1-> iota (canonS alpha i).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition const_pathS i :=
    clos_trans_1n T1 (fun alpha beta => beta = canonS alpha i).

@@

  \begin{Coqsrc}
Theorem Theorem_2_4 (lambda : T1) :
   nf lambda ->
   limitb lambda  ->
   forall i j, (i < j)%nat ->
               const_pathS 0 (canonS lambda j)
                             (canonS lambda i). 
  \end{Coqsrc}

@@

\begin{Coqsrc}
Fixpoint gnaw (alpha : T1) (s: list nat) :=
  match s with
    | nil => alpha
    | (0::s') => gnaw  alpha s'
    | (S i :: s')  =>  gnaw (canonS i alpha) s'
  end.

@@

\begin{Coqsrc}
Compute pp (gnaw (omega * omega) (interval 6 70)).
\end{Coqsrc}

\begin{Coqanswer}
 = (omega * 2 + 56)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
Compute (gnaw (omega * omega) (interval 6 700)).
\end{Coqsrc}

@@

\vspace{4pt}
 \noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets_Examples.html}{Epsilon0.Large\_Sets\_Examples}}
\begin{Coqsrc}
Example Ex1 : mlarge (omega * omega) (interval 6 510).
\end{Coqsrc}

@@

\begin{Coqsrc}
Section lim.
  Variables (lambda : T1)
            (Hnf : nf lambda)
            (Hlim : limitb lambda)
            (f : nat -> nat -> nat)
            (H : forall k, L_spec (canonS lambda k) (f (S k))).

@@

Lemma L_omega_ok : L_spec omega L_omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma path_to_omega_mult (i k:nat) :
  path_to (omega * i) (interval (S k) (2 * (S k))) (omega * (S i)).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma omega_mult_mlarge_0 i  : forall k,
    mlarge  (omega * (S i))
            (interval (S k)
                      (Nat.pred (iterate (fun p =>  S (2 * p)%nat)
                                         (S i)
                                         (S k)))).
\end{Coqsrc}

@@

Lemma L_omega_mult_ok (i: nat) :  L_spec (omega * i) (L_omega_mult i).
\end{Coqsrc}

@@

Lemma L_omega_square_ok: L_spec (omega * omega) 
          L_omega_square.

@@

\begin{Coqsrc}
Notation  "'one'"  := [zero,zero] : T2_scope.

(** The (n+1)-th finite ordinal *)
Notation "'FS' n" := (gcons zero zero n zero) (at level 10) : T2_scope.

>>>>

\begin{itemize}
\item A ``mathematical'' representation of the set of countable ordinal numbers, afer Kurt Schütte~\cite{schutte}. This representation uses several (hopefully harmless) axioms. We use it as a reference for proving the correctness of ordinal notations.
\item A family of \emph{ordinal notations} (also called \emph{[ordinal] notation systems}), \emph{i.e.} data types used to represent segments $[0,\mu)$, where $\mu$ is some countable ordinal. Each ordinal notation is defined inside the Calculus of Inductive Constructions (without axioms). Many functions are defined, allowing proofs by computation. Note that proofs of 
correctness of a given ordinal notation with respect to Schütte's model obviously use axioms.
Please execute the \texttt{Print Assumptions} command in case of doubt.
\end{itemize}

@@

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as quite simple data types in \gallina.

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations.ON_Generic.html}{OrdinalNotations.ON\_Generic}}

@@

Notation omega := (inr  0:t).
\end{Coqsrc}

@@

Notation omega := (1,0).

@@

\begin{figure}[h]
   \centering
   \begin{tikzpicture}[very thick, scale=0.6]
\begin{scope}[color=blue]
\node (A) at (0,0) {$A$};
\node(A0) at (2,0)[label=below:$0$]{$\bullet$};
\node(A1) at (3,0)[label=below:$1$]{$\bullet$};
\node(A2) at (4,0)[label=below:$2$]{$\bullet$};
\node (Adots) at (6,0) {$\ldots$};
\end{scope}
\begin{scope}[color=red]
\node (B) at (0,2) {$B$};
\node(B0) at (2,2)[label=above:$0$]{$\bullet$};
\node(B1) at (3,2)[label=above:$1$]{$\bullet$};
\node(B2) at (4,2)[label=above:$2$]{$\bullet$};
\node (Bdots) at (6,2) {$\ldots$};
\node (b) at (8,2) [label=above:$\omega$]{$\bullet$};
\node (bsucc) at (9,2) [label=above:$\omega+1$]{$\bullet$};
\node (Bdots2) at (10,2) {$\ldots$};
\end{scope}
\begin{scope}[color=red!50!blue]
\draw [->,thin] (A0) -- node [auto] {$\iota$} (B0);
\draw [->,thin] (A1) -- node [auto] {$\iota$} (B1);
\draw [->,thin] (A2) -- node [auto] {$\iota$} (B2);
\draw [->,thin] (Adots) -- node [auto] {$\iota$} (Bdots);
\end{scope}
\end{tikzpicture}
   \caption{\textcolor{blue}{$A$} is a sub-segment  of \textcolor{red}{$B$}}
   \label{fig:subsegment}
 \end{figure}

@@

\begin{Coqsrc}
Notation one := (ocons zero 0 zero).

Notation FS n := (ocons zero n zero).
\end{Coqsrc}

@@

\label{sect:omega-notation2}
\begin{Coqsrc}
Notation omega := (ocons (ocons zero 0 zero) 0 zero).
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive ppT1 : Set :=
    P_fin : nat -> ppT1
  | P_add : ppT1 -> ppT1 -> ppT1
  | P_mult : ppT1 -> nat -> ppT1
  | P_exp : ppT1 -> ppT1 -> ppT1
  | P_omega : ppT1

Notation _omega := P_omega.

@@

\begin{Coqanswer}
     = (_omega ^ _omega + _omega ^ 3 * 5 + 2)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
Fixpoint compare (alpha beta:T1):comparison :=
  match alpha, beta with
    zero, zero => Eq
  | zero, ocons a' n' b' => Lt
  | _   , zero => Gt
  | (ocons a n b),(ocons a' n' b') =>
      (match compare a a' with 
       | Lt => Lt
       | Gt => Gt
       | Eq => (match Nat.compare n n' with
                | Eq => compare b b'
                | comp => comp
                end)
       end)
  end.
\end{Coqsrc}

@@

Notation omega  := _Omega.
\end{Coqsrc}

@@

\index{hydras}{Exercises} 
\begin{exercise}
  Prove that that the ordinal addition on \texttt{E0} is associative.

  \textbf{Warning:}  Please do not import theorems from
  \texttt{ordinals/Schutte/}, since that library uses axioms!
\end{exercise}

@@

\index{hydras}{Projects}
\begin{project}
The class \texttt{ON} of ordinal notations has been defined long after this 
chapter, and is not used in the development of the type \texttt{E0} yet.
A better integration of both notions should simplify the development on ordinals in Cantor normal form. This integration is planned for the future versions.

@@

\index{hydras}{Projects}
\begin{project}
  Prove that ordinal multiplication is left-distributive over ordinal addition; that is:
  $$\forall \alpha \, \beta \, \gamma < \epsilon_0, \alpha\times(\beta+\gamma) = \alpha\times\beta+ \alpha\times \gamma$$

\end{project}
\section{A variant for hydra battles}

@@

Our definition may look more complex than the mathematical one, but
uses plain structural recursion over the type \coqsimple{T1}. Thus, tactics like
\coqsimple{cbn}, \coqsimple{simpl}, \coqsimple{compute}, etc., are applicable.

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Canon.html\#canon}{Epsilon0.Canon}}

\label{Functions:canonS}
\label{Functions:canon}
\begin{Coqsrc}
  Fixpoint canon alpha (i:nat) : T1 :=
  match alpha with
    zero => zero
  | ocons zero 0 zero  => zero
  | ocons zero (S k) zero  => FS k
  | ocons gamma 0 zero => (match T1.pred gamma with
                            Some gamma' =>
                            match i with 0 => zero
                                    |   S j => ocons gamma' j zero
                            end
                          | None =>
                            ocons (canon gamma i) 0 zero
                           end)

  |  ocons gamma (S n) zero =>
     (match T1.pred gamma with
       Some gamma' =>
       (match i with
         0 =>  ocons gamma n zero
       | S j => ocons gamma n (ocons gamma' j zero)
       end)
       | None =>
            ocons gamma n (ocons (canon gamma  i) 0 zero)
      end)
   |  ocons alpha n beta => ocons alpha n (canon beta i)  
end.
\end{Coqsrc}

@@

\begin{Coqsrc}
Compute pp (canon (T1.omega ^ T1.omega) 3). 
\end{Coqsrc}

\begin{Coqanswer}
  = (_omega ^ 3)%pT1
     : ppT1
   \end{Coqanswer}

\begin{Coqsrc}
Compute pp (canon (T1.omega ^ T1.omega * 3) 42).  
\end{Coqsrc}

\begin{Coqanswer}
    = (_omega ^ _omega * 2 + _omega ^ 42)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
  Lemma canon_succ i alpha :
  nf alpha -> canon (succ alpha) i = alpha.
Proof.
 revert i; induction alpha.
 (* ... *)
\end{Coqsrc}

@@

\label{lemma:canonS_LT}
\begin{Coqsrc}
Lemma canonS_LT i alpha :
  nf alpha -> alpha <> zero ->
  canon alpha (S i) t1<  alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma canonS_limit_strong (lambda : T1) : 
     nf lambda ->
     limitb lambda  ->
     forall beta, beta t1< lambda ->
                  {i:nat | beta t1< canon lambda (S i)}.

@@

\section{Accessibility inside \texorpdfstring{$\epsilon_0$}{epsilon0} : paths}
\index{maths}{Ordinal numbers!Accessibility inside epsilon0}
\label{sect:pathes-intro}

@@

\begin{Coqsrc}
Definition transition_S i : relation T1 :=
fun alpha beta =>
    alpha <> zero /\ beta = canonS alpha (S i).

@@

\begin{Coqsrc}
Lemma canonS_iota i alpha :
    nf alpha -> alpha <> 0 ->
    iota alpha -1-> iota (canon alpha (S i)).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition const_pathS i :=
  clos_trans_1n T1 (fun alpha beta => alpha <> zero /\
                                      beta = canon alpha (S i)).

@@

\begin{Coqsrc}
Theorem KS_thm_2_4 (lambda : T1) :
   nf lambda ->
   limitb lambda  ->
   forall i j, (i < j)%nat ->
               const_pathS 0 (canon lambda (S j))
                             (canon lambda (S i)).
\end{Coqsrc}

@@

\begin{Coqsrc}
Fixpoint gnaw (alpha : T1) (s: list nat) :=
  match s with
    | nil => alpha
    | (0::s') => gnaw  alpha s'
    | (S i :: s')  =>  gnaw (canon alpha (S i)) s'
  end.

@@

\begin{Coqsrc}
Compute pp (gnaw (T1.omega * T1.omega) (interval 6 70)).
\end{Coqsrc}

\begin{Coqanswer}
 = (_omega * 2 + 56)%pT1
     : ppT1
\end{Coqanswer}

@@

\begin{Coqsrc}
Compute (gnaw (T1.omega * T1.omega) (interval 6 700)).
\end{Coqsrc}

@@

\vspace{4pt}
 \noindent
\emph{From Module~\href{../theories/html/hydras.Epsilon0.Large_Sets_Examples.html}{Epsilon0.Large\_Sets\_Examples}}
\begin{Coqsrc}
Example Ex1 : mlarge (T1.omega * T1.omega) (interval 6 510).
\end{Coqsrc}

@@

\begin{Coqsrc}
Section lim.
  Variables (lambda : T1)
            (Hnf : nf lambda)
            (Hlim : limitb lambda)
            (f : nat -> nat -> nat)
            (H : forall k, L_spec (canon lambda (S k)) (f (S k))).

@@

Lemma L_omega_ok : L_spec T1.omega L_omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma path_to_omega_mult (i k:nat) :
path_to (T1.omega * i) (interval (S k) (2 * (S k)))
            (T1.omega * (S i)).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma omega_mult_mlarge_0 i  : forall k,
    mlarge  (T1.omega * (S i))
            (interval (S k)
                      (Nat.pred (iterate (fun p =>  S (2 * p)%nat)
                                         (S i)
                                         (S k)))).
\end{Coqsrc}

@@

Lemma L_omega_mult_ok (i: nat) :  L_spec (T1.omega * i) (L_omega_mult i).
\end{Coqsrc}

@@

Lemma L_omega_square_ok: L_spec (T1.omega * T1.omega) 
          L_omega_square.

@@

\begin{Coqsrc}
Notation  one  := [zero,zero].

(** The (n+1)-th finite ordinal *)
Notation FS n := (gcons zero zero n zero).
